The curious world of constructive mathematics

### 2022-09-09 algar-lecture-notes.tex
Many more exercises

\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\FF}{\mathbb{F}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\G}{\mathcal{G}}
\let\acuteH\H
\newcommand{\konig}{K\acuteH onig}
\renewcommand{\H}{\mathcal{H}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\L}{\mathcal{L}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\GG}{\mathbb{G}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\ccc}{\mathfrak{c}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\nnn}{\mathfrak{n}}
\newcommand{\?}{\,{:}\,}
\renewcommand{\_}{\mathpunct{.}\,}

@@

\newcommand{\cl}{\mathrm{cl}}
\newcommand{\notnot}{\emph{not~not}}

@@

\begin{exercise}[An epistemic riddle on transcendental numbers]
Verify, using a proof by contradiction, that at least one of the numbers~$e +
\pi$ and~$e - \pi$ is transcendental; and that at least one of the
numbers~$e + \pi$ and~$e \cdot \pi$ is transcendental.
\end{exercise}

@@

{\noindent\scriptsize\emph{Note.} A set~$X$ is \emph{Kuratowski-finite} if and
only if, for some number~$n \in \NN$, there is a surjective map~$[n] \to X$,
where~$[n] = \{ 0,1,\ldots,n-1 \}$. More briefly, a set~$X$ is
Kuratowski-finite iff its elements can be enumerated: $X = \{x_1,\ldots,x_n \}$.\par}
\end{enumerate}
Show that the following statement implies Markov's principle (from
Exercise~\ref{ex:tautologies}):
\begin{enumerate}
\addtocounter{enumi}{3}
\item Every real number which is not zero is invertible.
\end{enumerate}
\noindent
Brouwerian counterexamples abound in constructive mathematics; when developing
a constructive account of a theory, they help to clearly demarcate its limits.
As such additional Brouwerian counterexamples can be found in most texts on
constructive mathematics, such
as~\cite{mines-richman-ruitenburg:constructive-algebra}; a compilation mainly
from constructive analysis can be found
in~\cite{mandelkern:brouwerian-counterexamples}.
\end{exercise}

@@

\begin{ex}Classical linear algebra teaches us that every finite system of
generators of a vector space over a residue field\footnote{Constructively, the
notion of a field bifurcates into several nonequivalent notions: A nontrivial
commutative ring with unit is a \ldots
\begin{enumerate}
\item a \emph{geometric field} or \emph{discrete field} iff
every element is zero or
invertible (as in~$\QQ$ or~$\overline{\QQ}$),
\item a \emph{residue field} iff
every element which is not invertible is zero (as in the Cauchy or Dedekind
reals), and
\item a \emph{field of fractions} iff
every element which is not zero is invertible (as in the localization of the
Cauchy or Dedekind reals at the set of non-zeros or in the \emph{generic local
ring} described in Lecture~\ref{lect:generic}).
\end{enumerate}These each have their
uses~\cite{johnstone:rings-fields-and-spectra,mulvey:repr}.}
can be thinned to a generating family in which no vector is redundant, hence to
a basis. By slightly massaging the result obtained by the double-negation
embedding, we obtain the constructive theorem that every finitely generated
vector space over a residue field does \emph{not not} possess a basis.

@@

@@

\vfill{\small%
\begin{verbatim}
- Existence of generic models
- Generic ring is a field
- Generic prime ideal
- Kripke–Joyal semantics
\end{verbatim}
}

>>>>

\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\FF}{\mathbb{F}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\CCC}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\G}{\mathcal{G}}
\let\acuteH\H
\newcommand{\konig}{K\acuteH onig}
\renewcommand{\H}{\mathcal{H}}
\newcommand{\OOO}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\L}{\mathcal{L}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\R}{\mathcal{R}}
\newcommand{\T}{\mathcal{T}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\renewcommand{\S}{\mathcal{S}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\V}{\mathcal{V}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\GG}{\mathbb{G}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\ccc}{\mathfrak{c}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\nnn}{\mathfrak{n}}
\newcommand{\?}{\,{:}\,}
\renewcommand{\_}{\mathpunct{.}\,}

@@

\newcommand{\cl}{\mathrm{cl}}
\newcommand{\notnot}{\emph{not~not}}
\newcommand{\Sh}{\mathrm{Sh}}

@@

\begin{exercise}[An epistemic riddle on transcendental numbers]
Verify, using a proof by contradiction, that at least one of the numbers~$e +
\pi$ and~$e - \pi$ is transcendental; and that at least one of the
numbers~$e + \pi$ and~$e \cdot \pi$ is transcendental.

{\noindent\scriptsize\emph{Note.} At time of writing, for none of these numbers
a (constructive or classical) proof of their transcendence is known.\par}
\end{exercise}

@@

{\noindent\scriptsize\emph{Note.} A set~$X$ is \emph{Kuratowski-finite} if and
only if, for some number~$n \in \NN$, there is a surjective map~$[n] \to X$,
where~$[n] = \{ 0,1,\ldots,n-1 \}$. More briefly, a set~$X$ is
Kuratowski-finite iff its elements can be enumerated: $X = \{x_1,\ldots,x_n \}$.\par}

\item Every subset of the (Cauchy or Dedekind) reals which is inhabited and
bounded from above has a supremum.

{\noindent\scriptsize\emph{Note.} A \emph{supremum} of a set~$M$ of reals is a
number~$s$ such that~$M \leq s$ (that is~$x \leq s$ for all~$x \in M$) and such
that for every number~$s'$ with~$M \leq s'$, $s \leq s'$. In constructive
mathematics, we can make finer distinctions between the classically equivalent
constructions of the real numbers: The reals constructed using Cauchy
sequences inject into the reals constructed using Dedekind cuts which in turn
inject into the MacNeille reals, and each serve a different purpose. The Cauchy
and Dedekind reals cannot constructively be shown to be complete in the sense
of this exercise, while the MacNeille reals can. Conversely, the rationals can
be shown to be dense in the first two kinds of reals but not in the MacNeille
reals~\cite[Section~D4.7]{johnstone:elephant}.\par}
\end{enumerate}
Show that the following statement implies Markov's principle (from
Exercise~\ref{ex:tautologies}):
\begin{enumerate}
\addtocounter{enumi}{4}
\item Every real number which is not zero is invertible.
\end{enumerate}
\noindent
Brouwerian counterexamples abound in constructive mathematics; when developing
a constructive account of a theory, they help to clearly demarcate its limits.
As such additional Brouwerian counterexamples can be found in most texts on
constructive mathematics, such
as~\cite{mines-richman-ruitenburg:constructive-algebra}; a compilation mainly
from constructive analysis can be found
in~\cite{mandelkern:brouwerian-counterexamples}.
\end{exercise}

@@

\begin{ex}Classical linear algebra teaches us that every finite system of
generators of a vector space over a residue field\footnote{\label{fn:fields}Constructively, the
notion of a field bifurcates into several nonequivalent notions: A nontrivial
commutative ring with unit is a \ldots
\begin{enumerate}
\item a \emph{geometric field} or \emph{discrete field} iff
every element is zero or
invertible (as in~$\QQ$ or~$\overline{\QQ}$),
\item a \emph{residue field} iff
every element which is not invertible is zero (as in the Cauchy or Dedekind
reals), and
\item a \emph{field of fractions} iff
every element which is not zero is invertible (as in the localization of the
Cauchy or Dedekind reals at the set of non-zeros or in the \emph{generic local
ring} described in Lecture~\ref{lect:generic}).
\end{enumerate}These each have their
uses~\cite{johnstone:rings-fields-and-spectra,mulvey:repr}.}
can be thinned to a generating family in which no vector is redundant, hence to
a basis. By slightly massaging the result obtained by the double-negation
embedding, we obtain the constructive theorem that every finitely generated
vector space over a residue field does \emph{not not} possess a basis.

@@

\begin{exercise}[{The sheaf of real functions as a
field~\cite[Exercise~28]{blechschmidt:generalized-spaces}}]
Let~$X$ be a topological space. Let~$\CCC$ be the sheaf of continuous real-valued
functions on~$X$.
\begin{enumerate}
\item Let~$U$ be an open of~$X$. Let~$f \in \CCC(U)$. Show that~$U \models (\exists g\?\CCC\_ fg =_\CCC
1)$ iff there is a function~$g \in \CCC(U)$ such that~$fg = 1$.
\item Show that~$\CCC$ is a \emph{residue field} in that
$X \models \forall f\?\CCC\_
  \bigl(\neg(\exists g\?\CCC\_ fg = 1)\bigr) \Rightarrow f = 0$.
\item Give an example of space~$X$ such that it is not the case that~$\CCC$ is a
field in the stronger sense that
$\Sh(X) \models \forall f\?\CCC\_ (\exists g\?\CCC\_ fg = 1) \vee f = 0$.
\item Let~$X = \CC$ and let~$\OOO$ be the sheaf of holomorphic functions on~$X$,
that is~$\OOO(U) = \{ f : U \to \CC \,|\, \text{$f$ holomorphic} \}$.
Show that, in classical mathematics, the sheaf~$\OOO$ is \emph{discrete} in that
\[ \Sh(\CC) \models \forall f\?\OOO\_ \forall g\?\OOO\_ f = g \vee \neg(f = g).
\]
\end{enumerate}

\end{exercise}
\begin{exercise}[{The geometric interpretation of double
negation~\cite[Exercise~26]{blechschmidt:generalized-spaces}}]
Let~$\varphi$ be a formula over an open~$U$ of a topological space~$X$.
\begin{enumerate}
\item Show that there is a largest open,
denoted~``$\llbracket\varphi\rrbracket$'', such
that~$\llbracket\varphi\rrbracket \models \varphi$.\smallskip

{\noindent\scriptsize\emph{Hint.} Consider the union of all opens~$V$ such that~$V
\models \varphi$.\par}

\item Show that~$X \models \neg\neg\varphi$ iff~$\llbracket\varphi\rrbracket$
is dense.

{\noindent\scriptsize\emph{Note.} A subset~$W \subseteq X$ is \emph{dense} iff
for every open~$T \subseteq X$, if~$W \cap T = \emptyset$ then~$T =
\emptyset$.\par}

\item Give a condition on the topological space~$X$ such that the principle of
excluded middle (or equivalently the principle of double negation elimination)
is valid in~$\Sh(X)$.
\end{enumerate}
\end{exercise}

\begin{exercise}[The fundamental properties of the sheaf semantics]
Complete the proof of Proposition~\ref{prop:sheaf-semantics}, so verify
monotonicity and locality of the sheaf semantics.
\end{exercise}

\begin{exercise}[{Unique local existence is global existence~\cite[Exercise~27]{blechschmidt:generalized-spaces}}]
Let~$F$ be a sheaf on a topological space~$X$. Let~$\varphi$ be a formula over~$X$. Assume
that~$\Sh(X) \models \exists!s\?F\_ \varphi(s)$, that is
\[ \Sh(X) \models \exists s\?F\_ \varphi(s) \quad\text{and}\quad
  \Sh(X) \models \forall s\?F\_ \forall t\?F\_ (\varphi(s) \wedge \varphi(t)
  \Rightarrow s = t). \]
Show that for any open~$U \subseteq X$, there is a unique section~$s \in F(U)$
such that~$U \models \varphi(s)$.

{\noindent\scriptsize\emph{Note.} If you do not know the general definition of
a sheaf, then consider only the specific sheaf~$\CCC$ of continuous real-valued
functions.\par}
\end{exercise}

@@

\begin{exercise}[Anonymously Noetherian rings]
A commutative ring~$A$ is \emph{anonymously Noetherian} if for every
ideal~$\aaa \subseteq A$ it is \notnot the case that~$\aaa$ is finitely
generated.
\begin{enumerate}
\item Show that residue fields in the sense of Footnote~\ref{fn:fields} are
anonymously Noetherian.
\item Verify that~$\ZZ$ is anonymously Noetherian.

{\noindent\scriptsize\emph{Note.} Either run a direct proof or use the
anonymous least number principle of Proposition~\ref{prop:minima}(a).\par}
\item Look up a proof of Hilbert's basis theorem, for instance the
proof in~\cite[Theorem~7.5]{atiyah-macdonald:commutative-algebra}, and check
that it can be massaged into a constructive proof of the statement that
polynomial rings over anonymously Noetherian rings are anonymously Noetherian.
\end{enumerate}
\end{exercise}

\begin{exercise}[{Prüfer domains as valuation domains~\cite[Exercise~43]{blechschmidt:generalized-spaces}}]
An \emph{integral domain} is a commutative ring such that~$1 \neq 0$ and such that~$xy = 0$
implies~$x = 0$ or~$y = 0$. A \emph{valuation domain} is an integral domain
such that for any two elements, one divides the other. A \emph{Prüfer domain}
is an integral domain such that every finitely generated ideal~$\aaa$ is locally a
principal ideal (in the sense that there exists a partition~$1 = f_1 + \cdots +
f_n$ such that, for each index~$i$, the ideal~$\aaa[f_i^{-1}]$ is a principal
ideal in~$A[f_i^{-1}]$).
\begin{enumerate}
\item Let~$A$ be a ring. Show that~$A^\sim$ is an integral domain if~$A$ is.
Does the converse hold?
\item Let~$A$ be a valuation domain. Show that any matrix over~$A$ can be put
into diagonal form by elementary row and column operations.
\item Let~$A$ be an integral domain. Show that~$A$ is a Prüfer domain if and
only if~$A^\sim$ is a valuation domain.
\item Let~$A$ be a Prüfer domain. Show that any matrix over~$A$ can locally be
put into diagonal form by elementary row and column operations, by applying the
result of part~(b) to~$A^\sim$.
\end{enumerate}
\end{exercise}

\begin{exercise}[{A basic version of Kaplansky's theorem~\cite[Exercise~44]{blechschmidt:generalized-spaces}}]
\begin{enumerate}
\item Let~$A$ be a local ring. Let~$\aaa \subseteq A$ be a finitely generated
ideal such that~$\aaa^2 = \aaa$. Show that~$\aaa = (0)$ or~$\aaa = (1)$.\smallskip

{\noindent\scriptsize\emph{Hint.} Nakayama's lemma.\par}
\item Let~$A$ be a local ring. Let~$M \in A^{n \times n}$ be an idempotent
matrix. Verify that~$M$ is similar to a diagonal matrix with entries~$0$ and~$1$
by applying part~(a) to the ideals of~$k$-minors of~$M$. Deduce that the
cokernel of~$M$ is finite free.
\item Let~$A$ be an arbitrary ring. Let~$M \in A^{n \times n}$ be an idempotent
matrix. Show that the cokernel of~$M$ is finite locally free, by applying the
result of part~(b) to~$A^\sim$.
\item Verify, without using the law of excluded middle or that~$A$ is
Noetherian, that an~$A$-module~$M$ is finitely generated and projective if and
only if it is finite locally free.\smallskip

{\noindent\scriptsize\emph{Note.} A self-contained solution is given in
Ref.~\cite{blechschmidt:kaplansky}.\par}
\end{enumerate}
\end{exercise}
